---
title: "01-EDA"
author: "David West"
date: "9/12/2020"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

data <- ggplot2::diamonds

# Reordering color so it's from worst to best. Just like all the other ordered features.
data <- data %>% 
  mutate(
    color = ordered(color,
                    levels = c(
                      "J",
                      "I",
                      "H",
                      "G",
                      "F",
                      "E",
                      "D"
                    )))
```

# Introduction
The purpose of this file is to do some initial exploratory analysis of the diamonds data set. This data set is provided in the ggplot2 package.

# EDA
## First Look
A sample, summary, and first glimpse of the diamonds data set is provided below:
```{r}
head(data)

summary(data)

glimpse(data)
```
There are `r nrow(data)` rows of data. Each row is one observation / one diamond. There are `r ncol(data)` columns of data. Each column is a variable / feature. The features are:
```{r}
names(data)
```
price: price in US dollars (`r min(data$price)` to `r max(data$price)`) <br />
carat: weight of the diamond (`r min(data$carat)` to `r max(data$carat)`) <br />
cut: quality of the cut (in order of worst to best; `r levels(data$cut)`) <br />
color: diamond color (in order of worst to best; `r levels(data$color)`) <br />
clarity: a measurement of how clear the diamond is (in order of worst to best; `r levels(data$clarity)`) <br />
x: length in mm (`r min(data$x)` to `r max(data$x)`) <br />
y: width in mm (`r min(data$y)` to `r max(data$y)`) <br />
z: depth in mm (`r min(data$z)` to `r max(data$z)`) <br />
depth: total depth percentage = 2 * z / (x + y) (`r min(data$depth)` to `r max(data$depth)`) <br />
table: width of the top of a diamond relative to its widest point (`r min(data$table)` to `r max(data$table)`) <br />

# Conclusion

